Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-04-12 article.tex
Re-run

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{array}
\usepackage{booktabs}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{mathtools}
% \usepackage{float}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made from a phylogeny 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
  }
  \label{fig:example_2}
  \giovanni{There is a weird NA at the top of the figure.}
\end{figure}

@@

\richel{hiero!}

@@

\richel{
  I would enjoy being able to write ...
}
\begin{lstlisting}
  twinning_params <- create_twinning_params(
    tree_prior = create_bd_tree_prior(),
    ...
  )
)
\end{lstlisting}
\richel{
  ... as that would connect the tree priors more
}

We bundle again all the \verb;pirouette; arguments using \verb;create_pir_params;:

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, 
    when hand-picking a generative tree prior versus the control.
    Here, the 'twin' column shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_3}
\end{figure}

@@

To answer ERQ 2 we used an evidence-based model selection procedure to obtain 
an 
error distribution.
However, it is unclear to what extent this error is due to the mismatch between 
the tree priors (from generative and inference models) or due to stochasticity. 
As mentioned in ERQs 1 and 3, \verb;pirouette; provides a tool to evaluate this, using 
the twinning process. We now apply it to the same pipeline followed in ERQ 2.

@@

"What is the error made by BEAST2 from a phylogeny 
derived from an unknown speciation model,
when selecting the best inference model, 
in relation to the background error?"

@@

Here, as in ERQs 2 and 3, we start with a phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params,
  twinning_params = twinning_params
)
\end{lstlisting}
\richel{
  Refer to earlier listing for this call to \texttt{create\_pir\_params}
}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
errors <- pir_run(
  phylogeny = phylogeny,
  pir_params = pir_params
)
\end{lstlisting}
\richel{
  Refer to earlier listing for this call to \texttt{pir\_run}
}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_plot(errors)
\end{lstlisting}
\richel{
  Refer to earlier listing for this call to \texttt{pir\_run}
}

The output is shown in figure~\ref{fig:example_4}\giovanni{figure~\ref{fig:example_4} needs to be closer}. The alignment
generated for the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 (as 
shown in table \ref{tab:evidences_example_4}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 (as 
shown in table \ref{tab:evidences_example_4}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative 
    model is known, compared to candidates with similar inference models, in 
    relationship to the background error.
  }
  \label{fig:example_4}
  \giovanni{NA on the top.}
\end{figure}

@@

The output is shown in figure~\ref{fig:example_5}.
\giovanni{figure~\ref{fig:example_5} needs to be closer}
The alignment
generated from the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 \richel{again} (as 
shown in table \ref{tab:evidences_example_5}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 \richel{again} (as 
shown in table \ref{tab:evidences_example_5_twin}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the error made by BEAST2 from a phylogeny, 
    when the generative model is known, 
    compared to candidates with similar inference models, 
    in relationship to the background error.
  }
  \giovanni{na on the top-right area. labels on x superimpose each other. We should change the names to their shortened versions: e.g. relaxed\_log\_normal $\to$ rln}
  \label{fig:example_5}
\end{figure}

>>>>

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{array}
\usepackage{booktabs}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{mathtools}
\usepackage{float}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made from a phylogeny 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    ERQ 2: the inference error from
    when selecting the best inference model.
  }
  \label{fig:example_2}
\end{figure}

@@

@@

We combine all the parameters using \verb;create_pir_params;:

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    ERQ 3: the inference error made 
    when hand-picking a generative tree prior versus the control.
    Here, the 'twin' column shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_3}
\end{figure}

@@

When no inference model can be hand-picked, 
one can select a best model from a set,
as shown in ERQ2.
The obtained error distribution contains the error accumulated in the pipeline,
without quantifying to what extent this error is due to the mismatch between 
the generative tree prior $\mathit{P}$ and the selected inference model 
$\mathit{I}$.
As twinning (see section 'Twinning') shows us the baseline error,
we can quantify the contribution of the tree prior to the error.

@@

"What is the inference error from a phylogeny 
derived from an unknown speciation model,
when selecting the best inference model, 
in relation to the baseline error?"

@@

Here, as in ERQs 2 and 3, we start with a phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params,
  twinning_params = twinning_params
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
errors <- pir_run(
  phylogeny = phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_plot(errors)
\end{lstlisting}

The output is shown in figure~\ref{fig:example_4}. The alignment
generated for the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 (as 
shown in table \ref{tab:evidences_example_4}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 (as 
shown in table \ref{tab:evidences_example_4}).

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative 
    model is known, compared to candidates with similar inference models, in 
    relationship to the background error.
  }
  \label{fig:example_4}
\end{figure}

@@

The output is shown in figure~\ref{fig:example_5}.
The alignment
generated from the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 \richel{again} (as 
shown in table \ref{tab:evidences_example_5}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 \richel{again} (as 
shown in table \ref{tab:evidences_example_5_twin}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the error made by BEAST2 from a phylogeny, 
    when the generative model is known, 
    compared to candidates with similar inference models, 
    in relationship to the background error.
  }
  \label{fig:example_5}
\end{figure}
